Network Data Model and BerlinMOD Benchmark

### 2010-04-08 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Runtimes and database statistics completed and added.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{3.0?}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorise
the data models into data models for objects moving freely in space and data
models for moving objects that are constrained by a given network. For both
categories several different data models have been developed. Two of them, one
data model for objects moving freely in the 2D plane and one for network
constrained moving objects have been implemented in \secondo{} DBMS. Both
handle with the histories of moving objects. In this paper we present the
comparison of the capabilities of the both data models using the \bmodb{} and
propose a extension of the \bmodb{} with specialised challenges for network
constrained data models called BerlinMOD/Net. The extension enables us to
compare the capabilities of different network data models with respect to the
the specialised challenges of the network constrained data model.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorise
them into data models for objects moving freely in space and data models for
objects which movement is constrained by a given network. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
 space and \cite{1146465,956692,VazWolfNetMod} for spatio-temporal data objects
that are constrained by networks to name just a few.

For our experiments we choosed one example data model as a representative for
each of the both categories to compare the capabilities of the both different
data models. The category of data models for spatio-temporal objects moving
freely in space is represented by the data model presented in \cite{335426}.
And the category of network constrained data models is represented by the
network data model presented in \cite{1146465}. The advantage of this choice is,
 that both data models are implemented in the \secondo{} DBMS and use the same
 temporal representation. So we can exclude that the use of different DBMS or
temporal representations bias our results.

We used the \bmodb{} \cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is available in the \secondo{}
 DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
we use for our comparison. So we only have to transform the \bmodb{} spatial and
 spatio-temporal data types once in our network data model representation. This
avoids sources of errors and makes the correctness controll of the query results
 more easy.

In our experiments the network data model outperforms the data model of free
movement in space significantly. The total run time over all queries for the
network data model is half as long as the total run time for the data model of
free movement in the space. So the further developement of network data models
seems to be useful.

Network data models have their own challenges like network distance computing
which are not covered by the \bmodb{} yet. Therefore we propose a extension of
the \bmodb{} with a additional query set BerlinMOD/NET covering the specialised
challenges for spatio-temporal databases handling with network constrained moving
objects to enable us to compare the capabilities of different network data models
with the \bmodb{}.

We will present our work as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
 and \ref{sec:netdatamod}) we used for our comparision. In section
\ref{sec:bmodbNetDataMod} we present the setup for our experiments (\ref{sec:scenario})
 and describe our transformation of the \bmodb{} data and queries in the network
 data model representation (\ref{sec:Translation}) before we show the results of
our experiments (\ref{sec:results}). In section \ref{sec:newqueries} we present
 our extension of the \bmodb{} query set. We conclude the paper with a summation
 of our results in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
As mentioned before in the past other data models for free movement in the space \cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But they don't
provide a combination of different implemented and supported data types together
 with a existing benchmark like the actual \secondo{} DBMS. So it is self-evident
for us to use the \secondo{} DBMS in combination with the provided data types and
the \bmodb{} to compare the capabilities of the both different data models.

In the next subsections we give short reminders of the \secondo{} database system (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models (\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments. More
detailed information for each of them can be taken from the original papers.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
 single data models. Hence \secondo{} can be easily extended by the user
implementing so called \secondo{} algebra modules to introduce  new data types
and operations on this data types. The user also may define additional viewers
for the graphical user interface or write optimization rules or cost functions
to extend the optimizer. \secondo{} is free available in the web \cite{secondoweb}
and comes with a number of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the space \ref{sec:bmodbdatamod} and the network data model \ref{sec:netdatamod}. Furthermore the \bmodb{} descripted in \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generation are implemented for the \secondo{} DBMS.
It is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of different moving object
data representations. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin \cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin.
This makes it possible to use the data set of the \bmodb{} for network constrained
 data models. Every moving object has a home node and a work node and every weekday
 there will be a trip from the home node to the work node in the morning and trip
from the work node back to the home node in the afternoon. In the evening and at
the weekend randomly chosen cars spend additional trips (one in the evening and up
to six at the weekend) to different randomly chosen targets. The number of observed
cars and the duration of the observation period can be influenced by the \bmodb{}
user by setting the $SCALEFACTOR$ to different values in the data generation
script. For example at $SCALEFACTOR$ 1.0 the data generator will create 2000
moving objects observed for 28 days. Each of them sending a GPS-signal every 2
seconds. This simulated signals are simplified so that time intervals when a car
 doesn't move or moves in the same direction at the same speed are merged into
one moving step. E.g. if the car holds 8 hours in front of the work node there
 will be only one entry in the cars history of movement with a time interval of
8 hours instead of 14.400 entries one for each GPS interval.

@@

In the TBA we have two relations. One of them (dataMcar) contains the static data
 for each object like licence, type, and model together with an object identifier.
The other relation (dataMtrip) contains for each object identifier several tuples
each of them containing a single trip of the moving object (e.g. each time the car
drives from home node to work node) or a longer stop (e.g. the time the car holds
in front of the office).

Besides the moving objects the \bmodb{} provides sets of $QueryPoints$,
$QueryRegions$, $QueryInstants$, $QueryPeriods$, and $QueryLicences$,
 each of them containing 100 pseudo randomly generated objects (points, regions,
time instants, time intervals, and licences) used in the benchmark queries.

For this data objects the \bmodb{} provides two sets of queries. One set addresses
range queries (BerlinMOD/R) and the  other one nearest neighbour queries(BerlinMOD/NN).
 In this paper we focus on the range queries, which are the main aspect of the
\bmodb{} up to now.

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects object identity (known / unknown),
dimension (standard / spatial / temporal / spatio-temporal),
query interval (point / range / unbounded),
condition type (single object / object relations),
and aggregation (with or without aggregation).

We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translations.
\subsection{BerlinMOD Data Model}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
2D space presented in \cite{594784,335426,352963}. All spatial positions are
given in x,y-coordinates. A single spatial position is represented by the data
type \dt{point}. A \dt{point} consists of a pair of \dt{real} values interpreted
as x,y-coordinates in the 2D plane. Streets are represented by \dt{line} values.
 A \dt{line} value consists of a set of half segments representing the geometry
of the line in the 2D plane. Each half segment consists of two \dt{point} values
which are interpreted as start and end point of the half segment. Regions are
represented by the data type \dt{region}. A \dt{region} consists of a set of
half segments interpreted as outer (and inner) border of the region in the 2D plane.

All this spatial data types and many standard data types can be lifted to become
time dependent \dt{moving} values. For all data types \dt{$\alpha$} the constructor
\dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$}) (short form \dt{m$\alpha$}).
 A car may be represented by a \dt{mpoint}. A \dt{mpoint} is a \dt{point} changing
its position within time. Therefore a \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first point represents the position of the
\dt{mpoint} at the start of the time interval and the second point represents the
position of the \dt{mpoint} at the end of the time interval. The \dt{point} is
assumed to move on the straight line between this two points with constant speed.
The speed is given by the ratio from the distance of the two points and the length
of the time interval of the unit. All units of a \dt{mpoint} must have disjoint
time intervals, because a car cannot be at two different positions at the same time.
The units are sorted by ascending time intervals. This spatio-temporal data model of \dt{moving} allows us to compute the position of a \dt{mpoint} at every time instant
within its definition time. We can also compute the time instant the point passed a
given position assumed the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by a \dt{intime}(\dt{point}) (short form \dt{ipoint}). A \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabular}{|l|p{9.5cm}|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected time intervals.\\
\hline
\dt{mbool} & A time dependent boolean value. The value within each \dt{ubool} will
 be constant \true{} or \false{} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabular}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a network and every position is given related
to this network. The data type \dt{network} is the central data type in the network
data model. All other data types of the network data model are related to a
\dt{network} by the unique network identifier that is part of each \dt{network}
object.

The \dt{network} object contains all spatial information of the represented network
in three main relations. The first relation contains the attributes of the routes
(streets) like id, route curve, route length, and two boolean flags indicating
if the route starts at the lexicographic smaller end point and if the lanes of
the route are separated like on German Highways or not. The second relation
contains all attributes of the junctions (street crossings) like the identifiers
of the first and second route crossing in the junction, the distance of the
junction from the start of the first respectively second route, tuple identifiers
 of the both routes in the routes relation, tuple identifiers of the sections
connected by this junction in the sections relation, and a connectivity code
telling us which lanes of the two routes are connected by the junction. The third
main relation of the \dt{network} object is the sections relation containing the
attributes of the sections (street parts between two junctions or a junction and
the end of the street) like the route identifier of the route the section belongs to,
 the tuple identifier of this route in the routes relation, start and end position
 of the section on the route, section curve, and two boolean flags $startssmaller$
and $dual$ with the same meaning as in the routes relation.

@@

Single positions in the network are given as \dt{gpoint} values. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a \dt{side} value
({$up$, $down$, $none$).

The \dt{side} value is necessary if the route is dual, it tells us if the position
is reachable form the $up$ or the $down$ \dt{side} of the route. For simple
routes or positions which are reachable form both sides of the route the \dt{side}
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of \dt{RouteIntervals}, and two boolean flags telling us if the \dt{gline} is defined
and if the set of \dt{RouteIntervals} is sorted.

Each \dt{RouteInterval} consists of a route identifier identifying the route
the \dt{RouteInterval} belongs to, and the start and the end position from the
\dt{RouteInterval} on this route\footnote{In the original paper the
 \dt{RouteInterval}
includes a additional parameter \dt{side} like the \dt{gpoint} does. But this
parameter is not part of the implementation yet.}. We call a \label{sec:sortedgline}
set of \dt{RouteIntervals} sorted if the following conditions are fullfilled:
\begin{enumerate}
	\item all \dt{RouteIntervals} are disjoint
	\item the \dt{RouteIntervals} are stored in ascending order of their route identifiers
	\item if two disjoint \dt{RouteIntervals} have the same route identifier the \dt{RouteInterval} with the smaller start position is stored first
	\item for all $RouteIntervals$ in the set the condition: $startPosition \le endPosition$ holds
\end{enumerate}
We introduced the sorted \dt{gline} because many algorithms take profit from sorted
\dt{gline} values. For example the computation if a \dt{gpoint} is inside a
\dt{gline} can be done in O($n$) for unsorted and O($\log n$) time for
sorted \dt{gline} values, if $n$ is the number of \dt{RouteIntervals} in the
\dt{gline}.

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline}
value represents a path between two \dt{gpoint} in the network, we need the
\dt{RouteIntervals} exactly in the sequence they are used in the path. This will
nearly never be a sorted set like defined before. We solved this dilemma by
introducing the sorted flag. Every algorithm which can take profit from a sorted
\dt{gline} checks this flag and uses the corresponding code. We store \dt{gline}
values sorted whenever this is possible to support faster query execution.

@@

In deviation from the original network data model we extended the implementation of the \dt{mgpoint} with four additional attributes:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of \dt{RouteIntervals} representing the positions ever
traversed by the \dt{mgpoint}
	\item A boolean defined flag for the set of \dt{RouteIntervals}
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of \dt{RouteIntervals} was introduced, because it makes it much
faster to decide if a \dt{mgpoint} ever passed a given place or not. Instead of
 a linear check of all $m$ \dt{ugpoint}s of a \dt{mgpoint} we can perform a
binary scan on the much lower number $r$ of \dt{RouteIntervals}. This reduces
the time complexity from O($m$) to O($\log r$) for all \op{passes}
operations. Logically this should be done by a sorted \dt{gline} value but the
\secondo{} DBMS restricts us to use a sorted set of \dt{RouteIntervals} instead.

>>>>

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[boxed]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption,subcaption}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
\newcommand{\secver}{2.9.StableNewFlob}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past several data models for the representation of spatio-temporal data
objects have been developed. Among others we can categorize this data models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In the mean time two representants
of this data models, one for each category, have been implemented in
\secondo{} DBMS. We used the \bmodb{} to compare the capabilities of the both
data models. The results are shown in this paper.

The very good results of the network data model in this comparsion encourages us
to propose a extension of the \bmodb{} with a additional query set BerlinMOD/Net.
This new query set covers the specialised challenges for network constrained
data models and enables us to use the \bmodb{} to compare the capabilities of
different network data models with respect to this specialised network challenges.
\end{abstract}
\section{Introduction}
In the past several data models for the representation of spatial and
spatio-temporal data objects have been developed. Among others we can categorize
them into data models for objects moving freely in (two dimensional) space and
data models for network constrained moving objects. For both categories
several different data models have been presented like
\cite{335426,chenzaniolosqlst} for spatio-temporal data objects moving freely in
 space and \cite{1146465,956692,VazWolfNetMod} for spatio-temporal data objects
that are constrained by given networks to name just a few.

For our experiments we choosed one example data model as representative for
each of the both categories to compare the capabilities of the both different
data models. Spatio-temporal objects moving freely in two dimensional space are
represented by the data model presented in \cite{335426}. And network constrained
data models are represented by the network data model presented in \cite{1146465}.
Both data models are available in the \secondo{} DBMS and use the same temporal representation. So we can exclude that the use of different DBMS or temporal
representations bias the results of our comparision.

As we use \secondo{} it seems to be a logical next step to use the \bmodb{} \cite{BerlinMODVLDB} to compare the capabilities of the
both data models. The \bmodb{} is best to our knowledge the first benchmark for
complete spatio-temporal database systems and it is also available in the \secondo{}
 DBMS. Furthermore the \bmodb{} data model is the data model of freely movement
in two dimensional space that we use for our comparison. So we only have to
transform the \bmodb{} spatial and spatio-temporal data types once into our
network data model representation. This avoids sources of errors and makes the
correctness controll of the query results more easy.

In our experiments the network data model outperforms the data model of free
movement in space significantly. The total run time for all benchmark queries
is less than half as long for the network model as the total run time for
the data model of free movement in the space. Such that it seems useful to
develope specialised databases and operations for network constrained moving
objects. In the same way the storage space used for the trips in the network
data model is less than 60\% of the storage space used by the data model of free
movement in the space. Also the number of units in the network data model is
less than 50\% of the number of units in the data model of free movement in the
space.

The very good results of the network data model encourages us to propose the extension
of the \bmodb{} by a query set BerlinMOD/Net. The new query set covers the specialised
challenges of network data models like network distance computing which are not
covered by the original \bmodb{}. The new query set should enable us to compare
the capabilities of different network data models with respect to the special
challenges of network constrained data models.

The rest of the paper is organised as follows: In section \ref{sec:relWork} we give a
short reminder of the underlying \secondo{} DBMS (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}) and the two data models (\ref{sec:bmodbdatamod}
 and \ref{sec:netdatamod}) we choosed as representatives for our comparision.
In section \ref{sec:bmodbNetDataMod} we present the setup for our experiments (\ref{sec:scenario}) and describe our transformation of the \bmodb{} data and
queries in the network data model representation (\ref{sec:Translation}). The
results of our experiments are presented in section (\ref{sec:results}). At least
we present our proposed extension of the \bmodb{} for network data models in section \ref{sec:newqueries} and conclude our work in section \ref{sec:summary}.
\section{Related Work}
\label{sec:relWork}
As mentioned before in the past other data models for free movement in the space \cite{335426,chenzaniolosqlst} and for network constrained movement have been
presented \cite{1146465,956692,VazWolfNetMod}. As well there are some more
benchmarks \cite{COSTBenchmark, QueriesTheodoridis} and database systems for
spatial and spatio-temporal datatypes \cite{HERMES,1054151}. But they don't
provide a combination of different implemented and supported data types together
 with a existing benchmark like the actual \secondo{} DBMS. So it is self-evident
for us to use the \secondo{} DBMS in combination with the provided data types and
the \bmodb{} to compare the capabilities of the both different data models.

In the next subsections we give short reminders of the \secondo{} database system (\ref{sec:secondo}), the \bmodb{} (\ref{sec:bmodb}), and the both data models (\ref{sec:bmodbdatamod}, \ref{sec:netdatamod}) we used in our experiments.
\subsection{Secondo}
\label{sec:secondo}
The extensible \secondo{} DBMS presented in \cite{686903,1054151} provides a
platform for implementing various kinds of data models. It provides a clean
interface between the data model independent system frame and the content of the
 single data models. Hence \secondo{} can be easily extended by the user
implementing so called \secondo{} algebra modules to introduce  new data types
and operations on this data types. The user also may define additional viewers
for the graphical user interface or write optimization rules or cost functions
to extend the optimizer. \secondo{} is free available in the web \cite{secondoweb}
and comes with a number of already implemented spatial and spatio-temporal data types
and operations including the spatio-temporal data model of free movement in the space \ref{sec:bmodbdatamod} and the network data model \ref{sec:netdatamod}. Furthermore the \bmodb{} descripted in \ref{sec:bmodb} has been developed in the \secondo{} DBMS.
For our experiments we used the \secondo{} version \secver{}.
\subsection{BerlinMOD Benchmark}
\label{sec:bmodb}
The \bmodb{} was presented in \cite{BerlinMODVLDB} \nocite{BerlinMOD} and the
provided scripts for the data generator are implemented as \secondo{} DBMS operations.
The \bmodb{} is available in the web \cite{berlinmodweb} and provides a well defined
data-set and queries for the experimental evaluation of different moving object
data representations. The \bmodb{} emphasises the development of complete systems
and simplifies experimental repeatability pointing out the weakness and the potency
of the benchmarked systems.

The data-sets of the \bmodb{} are created using the street map of the German
captial Berlin \cite{bbike} and statistical data about the regions of Berlin
\cite{bevberlin,berlinstadtatlas} as input relations.
The created moving objects represent cars driving in the streets of Berlin,
simulating the behaviour of people living and working in Berlin.
Every moving object has a home node and a work node. Every weekday each car will
do a trip from the home node to work the work node in the morning and vice versa
in the late afternoon. Besides this randomly choosen cars will make additional
trips in the evening and up to six times at the weekend to randomly choosen
targets in Berlin. Because for all this the street network of Berlin is used to
generate the moving object data sets the \bmodb{} can also be used to benchmark
network constrained moving object data models.

The number of observed cars and the duration of the observation period can be
influenced setting the $SCALEFACTOR$ to different values in the data generation
script of the \bmodb{}. For example at $SCALEFACTOR$ 1.0 the data generator
creates 2000 moving objects observed for 28 days. Each of them sending a
GPS-signal every 2 seconds. This simulated signals are simplified such that time
intervals when a car doesn't move or moves in the same direction at the same
speed are merged into one moving unit. This means that if for example a car parks
8 hours in front of the work node there will be only one entry in the cars history of movement with a time interval of 8 hours instead of 14.400 entries one for each
GPS interval.

@@

In the TBA we have two relations. One of them (dataMcar) contains the static data
for each object like licence, type, and model together with an object identifier.
The other relation (dataMtrip) contains for each object identifier several tuples
each of them containing a single trip of the moving object (e.g. each time the car
drives from home node to work node) or a longer stop (e.g. the time the car parks
in front of the office).

Besides the moving objects the \bmodb{} provides sets of $QueryPoints$,
$QueryRegions$, $QueryInstants$, $QueryPeriods$, and $QueryLicences$,
each of them containing 100 pseudo randomly generated data objects (points, regions,
time instants, time intervals, and licences) used in the benchmark queries.

The \bmodb{} provides two sets of queries. One set BerlinMOD/R addresses
range queries and the  other one BerlinMOD/NN nearest neighbour queries.
We will focus on the range queries in this paper, which are the main aspect of the
\bmodb{} up to now.

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item object identity (known / unknown),
  \item dimension (standard / spatial / temporal / spatio-temporal),
  \item query interval (point / range / unbounded),
  \item condition type (single object / object relations),
  \item aggregation (with or without aggregation).
\end{itemize}
We will present the 17 queries in more detail in section \ref{sec:queries}
together with our network data model translation of this queries.
\subsection{BerlinMOD Data Model}
\label{sec:bmodbdatamod}
The data model used by the \bmodb{} is the same data model of freely moving in
two dimensional space presented in \cite{594784,335426,352963}. All spatial
positions are given in x,y-coordinates. A single spatial position is represented
by the data type \dt{point}. A \dt{point} consists of a pair of \dt{real} values
interpreted as x,y-coordinates in the two dimensional space.
Streets are represented by \dt{line} values.
A \dt{line} value consists of a set of half segments representing the geometry
of the line in the two dimensional space. Each half segment consists of two
\dt{point} values
which are interpreted as start and end point of the half segment. Regions are
represented by the data type \dt{region}. A \dt{region} consists of a set of
half segments interpreted as outer (and inner) border of the region in the two
dimensional space.

All this spatial data types and many standard data types can be lifted to become
time dependent \dt{moving} values. For all data types \dt{$\alpha$} the constructor
\dt{moving} creates a new data type \dt{moving}(\dt{$\alpha$}) (short form \dt{m$\alpha$}).
A car may be represented by a \dt{mpoint}. A \dt{mpoint} is a point object changing
its position within time. Therefore a \dt{mpoint} consist of a set of units called
\dt{unit}(\dt{point}) (short form \dt{upoint}). Each \dt{upoint} consists of a time
interval and two \dt{point} values. The first point represents the position of the
\dt{mpoint} at the start of the time interval and the second point represents the
position of the \dt{mpoint} at the end of the time interval. It is assumed that
the point moves on the straight line between this two points with constant speed.
The speed is given by the ratio from the distance of the two points and the length
of the time interval of the unit. All units of a \dt{mpoint} must have disjoint
time intervals, because a car cannot be at two different positions at the same time.
The units are sorted by ascending time intervals. This spatio-temporal data model of \dt{moving} allows us to compute the position of a \dt{mpoint} at every time instant
within its definition time. We can also compute the time instant the point passed a
given position assumed the \dt{mpoint} ever passes this position. The position of a
\dt{point} at a given time instant is represented by a \dt{intime}(\dt{point}) (short form \dt{ipoint}). A \dt{ipoint} consists of a time instant and a \dt{point} value.

Some other data types of \secondo{} which are used in the \bmodb{} are shown in table \ref{tab:bmodbdatatypes}.
\begin{table}[H]
\begin{center}
\begin{scriptsize}
\begin{tabularx}{\textwidth}{|l|X|}
\hline
\textbf{Data Type} & \textbf{Description} \\
\hline
\dt{bool} & Usual boolean data type.\\
\hline
\dt{int} & Usual integer number.\\
\hline
\dt{real} & Usual real number.\\
\hline
\dt{instant} & A point in time.\\
\hline
\dt{periods} & A set of disjoint and not connected time intervals.\\
\hline
\dt{mbool} & A time dependent boolean value. The value within each \dt{ubool} will
 be constant \true{} or \false{} \\
\hline
\dt{mreal} & Time dependent real number. Each unit will be defined by a function
of time representing the \dt{real} value at each time instant.\\
\hline
\end{tabularx}
\end{scriptsize}
\caption{Other Data Types of \bmodb{}}
\label{tab:bmodbdatatypes}
\end{center}
\end{table}
\subsection{Network Data Model}
\label{sec:netdatamod}
The central idea of the network data model presented in \cite{1146465} is that
every movement is constrained by a network and every position is given related
to this network. The data type \dt{network} is the central data type in the network
data model. All other data types of the network data model are related to a
given \dt{network} object by the unique network identifier that is part of
each \dt{network} object.

The \dt{network} object contains all spatial information of the represented network
in three main relations. The first relation contains the attributes of the routes
(streets) like id, route curve, route length, and two boolean flags indicating
if the route starts at the lexicographically smaller end point or not and if the
lanes of the route are separated like on German Highways or not.
The second relation
contains all attributes of the junctions (street crossings) like the identifiers
of the first and second route crossing in the junction, the distance of the
junction from the start of the first respectively second route, tuple identifiers
of the both routes in the routes relation, tuple identifiers of the sections
connected by this junction in the sections relation, and a connectivity code
telling us which lanes of the two routes are connected by the junction. The third
main relation of the \dt{network} object is the sections relation containing the
attributes of the sections (street parts between two junctions or a junction and
the end of the street) like the route identifier of the route the section belongs to,
the tuple identifier of this route in the routes relation, start and end position
of the section on the route, section curve, and two boolean flags $startssmaller$
and $dual$ with the same meaning as in the routes relation.

@@

Single positions in the network are given as \dt{gpoint} values. Besides the
network identifier a \dt{gpoint} consists of a route identifier, a distance from
the start of the route to the position of the \dt{gpoint} and a $side$ value
({$up$, $down$, $none$). The $side$ value is necessary if the route is dual.
It tells us if the position
is reachable from the $up$ or the $down$ side of the route. For simple
routes or positions which are reachable from both sides of the route the side
value is always $none$.

Parts of the network, regardless if they represent paths or regions, are given as
\dt{gline} values. Besides the network identifier a \dt{gline} consists of a set
of route intervals, and two boolean flags telling us if the \dt{gline} is defined
and if the set of route intervals is sorted.

Each route interval consists of a route identifier identifying the route
the route interval belongs to, and the start and the end position from the
route interval on this route
\footnote{In the original paper the
route interval includes a additional parameter $side$ analogous to the \dt{gpoint}.
But this parameter is not part of the implementation yet.}.
We call a \label{sec:sortedgline}
set of route intervals sorted if the following conditions are fullfilled:
\begin{itemize}
	\item all route intervals are disjoint
	\item the route intervals are stored in ascending order of their route identifiers
	\item if two disjoint route intervals have the same route identifier the route interval
  with the smaller start position is stored first
	\item for all route intervals in the set the condition: $startPosition \le endPosition$ holds
\end{itemize}
We introduced the sorted \dt{gline} because many algorithms take profit from sorted
\dt{gline} values. For example the computation if a \dt{gpoint} is inside a
\dt{gline} needs O($n$) time for unsorted and O($\log n$) time for
sorted \dt{gline} values, if $n$ is the number of the route intervals in the
\dt{gline}.

Unfortunately not all \dt{gline} values can be stored sorted. If a \dt{gline}
value represents a path between two \dt{gpoint} in the network, we need the
route intervals exactly in the sequence they are used in the path. This will
nearly never be a sorted set like defined before. We solved this dilemma by
introducing the boolean sorted flag. Every algorithm which can take profit from a sorted
\dt{gline} checks this flag and uses the corresponding code. We store \dt{gline}
values sorted whenever this is possible to support faster query execution.

@@

In deviation from the original network data model we extended the implementation
of the \dt{mgpoint} with four additional attributes:
\begin{enumerate}
	\item The total driven distance
	\item A sorted set of route intervals representing the positions ever
traversed by the \dt{mgpoint}
	\item A boolean defined flag for the set of route intervals
	\item A spatio-temporal minimum bounding box
\end{enumerate}
The sorted set of route intervals was introduced, because it makes it much
faster to decide if a \dt{mgpoint} ever passed a given network position or not.
Instead of a linear check of all $m$ \dt{ugpoint}s of a \dt{mgpoint} we can
perform a binary scan on the much lower number $r$ of route intervals. This reduces
the time complexity from O($m$) to O($\log r$) for all \op{passes}
operations. Logically this should be done by a sorted \dt{gline} value but the
\secondo{} DBMS restricts us to use a sorted set of route intervals instead.

The spatio-temporal minimum bounding box was introduced as parameter to the
\dt{mgpoint} because the computation of this value is very expensive in the
network data model. Although each unit of a \dt{mgpoint} stays on the same route
at same speed the motion may follow different spatial directions, e.g. a route may
lead uphill in serpentine. Not all this positions must be enclosed by a bounding box
computed just from the spatial position of the $start$ and the $end position$
of the unit. Therefore we have always to examine the spatial dimensions of the
complete part of the route passed within a unit to compute the units bounding box.
All spatial information of the route curve is hidden in the \dt{network} object.
We have to call the route curve from the \dt{network} object to compute the
spatial dimensions of the unit bounding box. If $r$ is the number of routes
of the network and $h$ the number of half segments of the traversed part of
the route curve passed in a unit we need O($h + \log r$) time to
compute the bounding box for a single unit. The bounding box of the \dt{mgpoint}
is the union of the bounding boxes of its $m$ units. So the computation of
a \dt{mgpoint} bounding box takes O($m(h + \log r)$) time. This is a
very expensive computation and the bounding box of a \dt{mgpoint} is only computed
on  demand or if we can get it for free. For example we can copy the bounding box
of a \dt{mpoint} if we translate it into a \dt{mgpoint} without computational effort.
But we don't maintain this attribute at every change of the \dt{mgpoint}. If the
\dt{mgpoint} changes we set the bounding box attribute to be undefined and compute
it again on demand if necessary.
\section{BerlinMOD and Network Data Model}
\label{sec:bmodbNetDataMod}
In the next subsections we first present our experimental setup (\ref{sec:scenario})
and the translations of \bmodb{} data-sets and queries into our network data model  representation (\ref{sec:Translation}) including the description of the indexes
we build to support faster query execution. We conclude this section with the
results of our experiments in \ref{sec:results}.
\subsection{Experimental Setup}
\label{sec:scenario}
For our experiments we used a Standard PC with a AMD Phenom II X4 Quad Core 2.95 GHz CPU,
8 GB main memory, and 2 TB HDD. We used the Linux openSUSE 11.2 operating system and
installed \secondo{} DBMS version \secver{} and the \bmodb{} from the web.
